### 2018-05-27 write-up/implementation.tex
Finish Conclusion

\prechapter{%
    I will describe the structure of LT4LA and explain the features of its core
    language. I will then show how typical linear-algebra programs can be
    elaborated into the core language and checked for linearity. Finally,
    I will explain how such programs can be translated (in this particular
    implementation) to OCaml code that is not obviously safe (with
    respect to aliasing, read/write permissions and memory management).
    Although I implemented LT4LA in OCaml, I believe the ideas described in
    this chapter can be applied easily to other languages and also are modular
    enough to extend the OCaml implementation to output to different back-end
    languages.
}%

\section{Structure of LT4LA}

LT4LA follows the structure of a typical compiler for a (E)DSL\@. From the
start, I made a concerted effort to (1) write pure-functional code (typically
using a monadic-style) which helped immensely with modularity and debugging
when tests showed errors (2) produce readable, useful and precise
error-messages in the hope that someone who did not understand linear types
could still use the LT4LA (3) write tests and set-up
continuous-integration for all non-trivial functions so that I could spot and
correct errors that were not caught by OCaml's type-system whenever I
implemented new features or or refactored my code.

@@

\section{Core Language}

A full description of the core language can be found in
Appendix~\ref{chap:ott_spec}. Its main features are intuitionistic values,
value-restriction, fractional-capabilities (inferrable at call sites),
if-expressions and recursion.

@@

To make a linearly-typed language usable, we need some way of using values zero
or more than once, as we would an intuitionistic value. For this, we have in
the type-expressions the !-constructor and in the term-expressions the
\ltfla{Many}-constructor and the \ltfla{let Many <id> = .. in .. } eliminator.
The idea behind the !-type is that the value uses no `resources'
(linearly-typed expressions). To start off with, it is enough to say that
anything which can be passed around by copying, will have a !-type. This
includes integers, elements and booleans. So \ltfla{3 : !int} and \ltfla{3. +.
4. : !elt}. However, all bindings are still linear by default, so to emulate
intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to  \ltfla{let Many x
= <exp> in let Many x = Many (Many x) in <body>} (similarly for function
argument bindings). The reader can check (using the rules in
Appendix~\ref{chap:ott_spec}) that this has the effect of moving \ltfla{x : !t}
from the linear to the intuitionistic environments, only if \ltfla{<exp> : !t}.

@@

Not quite, but to understand how we can fix this problem, we need to question
an assumption left implicit up until this point: what does \ltfla{Many} even
mean? What does it do at runtime? One option is to go down the C++ route and
make \ltfla{Many} act like a \ltfla{shared_ptr} and act as a runtime
reference-count for arrays. I chose to not go for this option because it went
against the \emph{explicitness} and \emph{predictability} that C and Fortran
have. It would make analysing when and what is allocated and freed more like
the higher-level languages I was trying to move away from.

My aim is to show linear types can be simple to understand and apply to linear
algebra, enough so that it can be grafted (in a limited way) on to existing
languages as a \emph{library}. In that spirit, the simplest thing that
\ltfla{Many} can mean at runtime is \emph{nothing}. This language construct is
translated into a standard OCaml language constructor of the form \ltfla{type
'a bang = Many of 'a [@@unboxed]}. The \emph{unboxed} annotation means
that the type and its constructor only exist for the purpose of type-checking
in OCaml; \emph{the runtime representation of values of type \ltfla{'a} and
\ltfla{'a bang} is exactly the same}.

With this understanding, our problem is that arrays and matrices are unlike
other values such as integers and elements because (under the OCaml hood)
calling a function with an array argument copies a \emph{pointer} to the array
rather than the array itself, instead of the \emph{value} itself. So, we can
start making a distinction, \emph{defining} elements, integers, booleans,
intuitionistic variables, units and lambda-expressions that capture no linear
variables as \emph{values} (since they cannot break referential-transparency)
and anything else (arrays, matrices, expressions which can be reduced, such as
function application and if-expressions) as not being `values'. If this sounds
familiar, it is because this is the same \emph{value-restriction} `trick' from
the world of polymorphic types applied to linearity instead. We then have the
rule that we can only use \ltfla{Many} on expressions that are defined to be
\emph{values} and \emph{use no linear variables}.

@@

Having started off with linearity and understanding how it helps a programmer
keep track of of memory allocation and deallocation, and then understood and
correctly implemented intuitionism for the only values we wish to be
intuitionistic, and, we now tackle the problem of how to implement safe
aliasing and read/write permissions.

Array and matrix types are parameterised by \emph{fractional-capabilities}.  A
fraction of 1 ($2^0$) represents complete ownership of a value; in particular,
this allows a programmer to write or free it. Creating an array gives you
ownership of it; the function \ltfla{array : !int --o z arr} (where \ltfla{z}
represents `0'). Once you have ownership of an array, you can free it:
\ltfla{free : z arr --o unit}.  Importantly, because a linear-value may only be
used once, the array just freed is \emph{out of scope} for following
expressions, preventing use-after-free.  Ownership also enables you to write to
the array: \ltfla{set : z arr --o !int --o !elt --o z arr} (the syntax
\ltfla{w[i] := j} is just sugar for \ltfla{set w i j}). Here, linearity
prevents accessing aliases which represented the array \emph{before} the
mutation.

Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
for a positive integer $k$) represents read-only access. So, the \ltfla{'x}
represents a natural number (either a zero \ltfla{z}, variable \ltfla{'x} or a
successor (+1) of a natural number). Hence, you can read from (index) any array
\ltfla{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \ltfla{let !v
<- w[i]} is just sugar for \ltfla{let (w, !v) = get _ w i}). In general, a
left-arrow \ltfla{<-} signifies transparent rebinding with returned values: it
means a program can \emph{appear} to use a variable multiple times, important
for keeping LT4LA usable and readable. The underscore is how a programmer tells
the compiler to automatically \emph{infer} the correct fractional-capability
based on the other arguments passed to the function. In conjunction with the
requirement that functions declarations need type-annotations for their
arguments, this allows a fractional-capability to be correctly inferred in any
program.

Fractions exist to provide both safe aliasing and read/write permissions, via
the primitives \ltfla{share : 'x . 'x arr --o 'x s arr * 'x s arr} and
\ltfla{unshare: 'x .  'x s arr --o 'x s arr --o 'x arr}.  For the former, two
arrays returned (which happen to just be the given array) can now only be read
from and not written to. If you want to write to this array, you must use the
latter to combine other read-only aliases until you are left with a value of
type \ltfla{z arr}, guaranteeing no other aliases exists.

Given this set-up, we now \emph{statically} have \emph{perfect} information
about aliasing and ownership of values in the program. We can only write to or
free an array only when we own it; ownership guarantees no other aliases exist
in scope at the point of usage. In Figure~\ref{fig:ltfla_kalman}, I show how
this perfect information can be used to write more natural-looking code using
value-semantic expressions which behave in precisely the way we intend it to.
Now the programmer need not resort to manually figuring out and inserting
\texttt{noalias} annotations and worrying about what variables can and cannot
be written to or freed; instead they can let the loyal and tireless compiler do
the heavy lifting.

@@

Because we do not know which way a condition will evaluate at run time, we must
guarantee the both branches use the same set of linear variables. Writing the
type-checker in a pure-functional monadic style paid off here because I could
now sandwich monadic values with state-adjustments either side of it. Given two
monadic values that represented type-checking two branches of an if-expression,
I could use the code in Figure~\ref{fig:same_resources} to easily save, reset
and compare the state either side of running those monadic values.

@@

A non-recursive function may be used more than once if it does not refer to any
linear variables from the surrounding scope. So, we can desugar something like
\ltfla{let x = 3 in let !f (y : !int) = x + y in <body>} to \ltfla{let x = 3 in
let Many f =  Many (fun y : !int -> x + y) in  <body>}. Recursion is
slightly more complicated: we can desugar the factorial function
(Figure~\ref{fig:ltfla_factorial}) to \ltfla{let Many f = fix (f, x : !int,
if (*..*) : !int) in <body> }. However, \ltfla{fix}, like \ltfla{Many},
must also not use any linear variables from its surrounding scope.

\section{Matrix Expressions}

We have now arrived at an extended application of linear types. I will show
how, we can apply the ability to automatically check aliasing, read/write
permissions, memory allocation, re-use and deallocation, to the domain of
matrix-expression compilation.

\begin{sidewaysfigure}
    \inputminted[linenos, fontsize=\footnotesize]{fortran}{kalman.f90}
    \caption{Kalman filter in Fortran 90.}\label{fig:fortran_kalman}
\end{sidewaysfigure}

In Figure~\ref{fig:fortran_kalman}, we see the difficulty of efficiently
implementing a \emph{Kalman filter}, a powerful set of equations applicable
to a wide variety of problems. From the comments, we see that every variable is
annotated with the step/matrix expression that it will hold at some point
during the computation (an equivalent alternative, say in C++, could be to have
a meaningful name for each step/matrix expression and manually annotate/keep
track of which names alias the same location).

In contrast, Figure~\ref{fig:ltfla_kalman}, offers the advantages of
\begin{itemize}
    \item aliasing: labelling each step with a different, more meaningful variable name,
    \item easily spotting which resources are being passed in and which are
        allocated for the function (new/copy),
    \item unambiguously seeing \emph{when} and what values are freed;
\end{itemize}
and have the compiler automatically ensure the safety of each of the above by respectively
\begin{itemize}
    \item making it impossible to refer to steps/values which are no longer usable,
    \item ensuring all values are declared and \emph{initialised} correctly before they are used,
    \item checking values are neither used after they are freed \emph{nor} leaked.
\end{itemize}

@@

\begin{figure}[tp]
    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/kalman.lt}
    \caption{Kalman filter in LT4LA.}\label{fig:ltfla_kalman}
\end{figure}

@@

All of the syntax in Figure~\ref{fig:ltfla_kalman} can be unambiguously
desugared \emph{before} type-checking, through fairly simple pattern matching.
Matrix expression compilation is well-trodden territory in
academia~\cite{rocklin_thesis, fabregat_thesis, gunnels_flame, linnea, taco}
but this is, to my knowledge, the first type-based approach to it.

@@

\begin{figure}[tp]
    \begin{minted}[fontsize=\small]{ocaml}
let x <- [| y |] in ..              ==> let (y,x) = copyM_to _ y x in ..
let x <- new [| y |] in ..          ==> let (y,x) = copyM _ y in ..
let x <- [| a*b + c |] in ..        ==> let x <- [| 1.*a*b + 1.*c |] in ..
let x <- [| i.*a*b + j.*c |] in ..  ==> let ((a,b), c) = (* BLAS *) in ..
let x <- new (m, n) [| a*b |] in .. ==> let c = matrix m n in
                                        let x <- [| 1.*a*b + 0.*c |] in ..
    \end{minted}
    \caption{Syntactic translations of matrix expressions to linearly-typed
        matrix functions. Further annotations on the matrix variables
        (\ltfla{sym} or \ltfla{^T}) determine which BLAS routine and what
        parameters to call it with.}\label{fig:mat_patterns}

@@

One of the first hurdles I encountered was compositionality: to compile $AB +
C$ it is typically better to use a BLAS routine directly rather than but first
compiling $A$, then $B$, then adding a call to multiply them, then $C$ and
finishing with a call to add the results.

@@

    \item \ltfla{let Many x = Many <exp> in <body>} with \ltfla{let x = <exp> in <body},

@@

The end result is visible in Figures~\ref{fig:ocaml_sumarray}
and~\ref{fig:ocaml_kalman}: OCaml code that is not obviously safe and correct
with respect to linearity, as promised at the start of the chapter.

\begin{figure}[tp]
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let rec f i n x0 row =
  if Prim.extract @@ Prim.eqI i n then (row, x0)
  else
    let row, x1 = Prim.get row i in
    f (Prim.addI i (Many 1)) n (Prim.addE x0 x1) row
in
f
    \end{minted}
    \caption{Recursive OCaml function for a summing an array, generated from
        Figure~\ref{fig:ltfla_sumarray}, passed through \texttt{ocamlformat} for
        presentation.}\label{fig:ocaml_sumarray}

@@

\begin{figure}[p]
    \begin{minted}[linenos, fontsize=\footnotesize]{ocaml}
let kalman sigma h mu r_1 data_1 =
  let h, _p_k_n_p_ = Prim.size_mat h in
  let k, n = _p_k_n_p_ in
  let sigma_h = Prim.matrix k n in
  let (sigma, h), sigma_h =
    Prim.symm (Many true) (Many 1.) sigma h (Many 0.) sigma_h
  in
  let (sigma_h, h), r_2 =
    Prim.gemm (Many 1.) (sigma_h, Many false) (h, Many true) (Many 1.) r_1
  in
  let (h, mu), data_2 =
    Prim.gemm (Many 1.) (h, Many false) (mu, Many false) (Many (-1.)) data_1
  in
  let h, new_h = Prim.copy_mat_to h sigma_h in
  let r_2, new_r = Prim.copy_mat r_2 in
  let chol_r, sol_h = Prim.posv new_r new_h in
  let chol_r, sol_data = Prim.potrs chol_r data_2 in
  let () = Prim.free_mat chol_r in
  let h_sol_h = Prim.matrix n n in
  let (h, sol_h), h_sol_h =
    Prim.gemm (Many 1.) (h, Many true) (sol_h, Many false) (Many 0.) h_sol_h
  in
  let () = Prim.free_mat sol_h in
  let h_sol_data = Prim.matrix n (Many 1) in
  let (h, sol_data), h_sol_data =
    Prim.gemm (Many 1.) (h, Many true) (sol_data, Many false) (Many 0.) h_sol_data
  in
  let mu, mu_copy = Prim.copy_mat mu in
  let (sigma, h_sol_data), new_mu =
    Prim.symm (Many false) (Many 1.) sigma h_sol_data (Many 1.) mu_copy
  in
  let () = Prim.free_mat h_sol_data in
  let h_sol_h_sigma = Prim.matrix n n in
  let (sigma, h_sol_h), h_sol_h_sigma =
    Prim.symm (Many true) (Many 1.) sigma h_sol_h (Many 0.) h_sol_h_sigma
  in
  let sigma, sigma_copy = Prim.copy_mat_to sigma h_sol_h in
  let (sigma, h_sol_h_sigma), new_sigma =
    Prim.symm (Many false) (Many (-1.)) sigma h_sol_h_sigma (Many 1.) sigma_copy
  in
  let () = Prim.free_mat h_sol_h_sigma in
  ((sigma, (h, (mu, (r_2, sol_data)))), (new_mu, new_sigma)) )
in
kalman
    \end{minted}
    \caption{OCaml code for a Kalman filter, generated from
        Figure~\ref{fig:ltfla_kalman}, passed through \texttt{ocamlformat} for
        presentation.}\label{fig:ocaml_kalman}

@@

Advantages of generating code at compile time include avoiding runtime overhead
as well as catching any interface changes between the generated code and the
code that uses it via OCaml's type system (another benefit of embedding LT4LA's
type system inside of OCaml).

I suspect that this approach will be very valuable to not just users of
libraries such as Numpy or Owl, but also their implementors: library functions
which presented a safe, value-semantic interface but use unsafe, mutating
operations on the inside could now be expressed using LT4LA and have an extra
safety-net and automatic checking for their implementations.

>>>>

\prechapter{%
    I will describe the structure of LT4LA: a DSL (domain specific language)
    \textbf{library} written in OCaml that transpiles to OCaml. I will explain
    the features of its core language and show how readable (equational and
    algebraic expressions) and easy to reason about linear-algebra programs
    can be elaborated into memory-efficient ones in the core language that are
    then checked for safety (with respect to aliasing, read/write permissions,
    memory allocation, re-use and de-allocation). Finally, I will explain how
    such programs can be transpiled to OCaml code that is not obviously safe.
    Although I refer to OCaml-specific features of the type-checker and
    transpiler, I believe the ideas described in this chapter can easily be
    implemented in other languages and are also general enough to transpile to
    different back-end languages, such as C or Fortran.
}%

\section{Structure of LT4LA}\label{sec:structure_lt4la}

LT4LA follows the structure of a typical compiler for a (E)DSL\@. From the
start, I made a concerted effort to (1) write pure-functional code (typically
using a monadic-style) which helped immensely with modularity and debugging
when tests showed errors (2) produce readable, useful and precise
error-messages in the hope that someone who did not understand linear types
could still use the LT4LA (3) write tests and set-up
continuous-integration for all non-trivial functions so that I could spot and
correct errors that were not caught by OCaml's type system whenever I
implemented new features or refactored my code.

@@

\section{Core Language}\label{sec:core_lang}

A full description of the core language can be found in
Appendix~\ref{chap:ott_spec}. For convenience, I have reproduced its type
grammar and defined how I interpret those types in OCaml's type system in
Figure~\ref{fig:type_grammar}.  The point of LT4LA is to keep the same values
and types we are familiar with, but \emph{restrict their usage} in certain
ways. LT4LA's core language's main features are: intuitionistic values,
value-restriction, fractional-capabilities (inferrable at call sites),
if-expressions and recursion.

\begin{figure}[tp]
    \centering
    \begin{minipage}{.3\textwidth}
        \centering
        \begin{grammar}
            <f> ::= `'
            \alt <fc>
            \alt `Z'
            \alt `S' <f>

            <t> ::= `'
            \alt `unit'
            \alt `bool'
            \alt `int'
            \alt `elt'
            \alt <f> `arr'
            \alt <f> `mat'
            \alt `!' <t>
            \alt \synt{$\forall$} <fc.> <t>
            \alt <t> \lit{$\otimes$} \synt{$t'$}
            \alt <t> \lit{$\multimap$} \synt{$t'$}
        \end{grammar}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \centering
        \begin{minted}[fontsize=\small, linenos]{ocaml}
module Arr =
  Owl.Dense.Ndarray.D

type z = Z
type 'a s = Succ

type 'a arr =
  A of Arr.arr
  [@@unboxed]

type 'a mat =
  M of Arr.arr
  [@@unboxed]

type 'a bang =
  Many of 'a
  [@@unboxed]
        \end{minted}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \begin{align*}
            [\![ f\!c ]\!] &= \texttt{'fc} \\
            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
            [\![ \textbf{int} ]\!] &= \texttt{int}\\
            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
        \end{align*}
    \end{minipage}
    \caption{Type grammar of LT4LA (left) and an informal interpretation of it
        into OCaml types (right).}\label{fig:type_grammar}
\end{figure}

@@

To make LT4LA a usable DSL, I needed some way of using values zero or more than
once, as a programmer would an intuitionistic value. For this, I added the
!-constructor at the type-level of the DSL and the \ltfla{Many}-constructor and
the \ltfla{let Many <id> = .. in .. } eliminator at the term-level of the DSL.
The idea behind the !-type is that the value uses no `resources'
(linearly-typed expressions). To start off with, it is enough to say that
anything which can be passed around by copying, will have a !-type. This
includes integers, elements and booleans. So \ltfla{3 : !int} and \ltfla{3. +.
4. : !elt}. However, all bindings are still linear by default, so to
emulate intuitionism, I desugar \ltfla{let !x = <exp> in <body>} to
\ltfla{let Many x = <exp> in let Many x = Many (Many x) in <body>}
(similarly for function argument bindings). The reader can check (using the
rules in Appendix~\ref{chap:ott_spec}) that this has the effect of moving
\ltfla{x : !t} from the linear to the intuitionistic environments, only if
\ltfla{<exp> : !t}.

@@

Not quite, but to understand how we can fix this problem, we need to question
an assumption left implicit up until this point: what does \ltfla{Many} even
mean in the DSL? How is it translated to OCaml? What is the OCaml runtime
behaviour of the DSL's \ltfla{Many}-constructor? One option is to go down the
C++ route and make \ltfla{Many} act like a \ltfla{shared_ptr} and act as a
runtime reference-count for arrays. I chose to not go for this option because
it went against the \emph{explicitness} and \emph{predictability} that C and
Fortran have. It would make analysing when and what is allocated and freed more
like the higher-level languages I was trying to move away from.

My aim is to show linear types are simple to understand and apply to
linear-algebra libraries, enough so that it can be grafted (in a limited way)
on to existing languages as a \emph{library}. In that spirit, the simplest
thing that the DSL's \ltfla{Many}-constructor can mean in the OCaml runtime is
\emph{nothing}.  This language construct is translated into a standard OCaml
type-and-constructor of the form \ltfla{type 'a bang = Many of 'a [@@unboxed]}.
The \emph{unboxed} annotation means that the type and its constructor only
exist for the purpose of type-checking in OCaml; \emph{the OCaml runtime
representation of values of type \ltfla{'a} and \ltfla{'a bang} is exactly the
same}.

With this understanding, our problem is that arrays and matrices are unlike
other values such as integers and elements because (under the OCaml hood)
calling a function with an array argument copies a \emph{pointer} to the array
rather than the array itself, instead of the \emph{value} itself. So, we can
start making a distinction, \emph{defining} elements, integers, booleans,
intuitionistic variables, units and lambda-expressions that capture no linear
variables as \emph{values} (since they cannot break referential-transparency)
and anything else (arrays, matrices, expressions which can be reduced, such as
function application and if-expressions) as not being `values'. If this sounds
familiar, it is because this is the same \emph{value-restriction} `trick' from
the world of polymorphic types applied to linearity instead. We then have the
rule that in LT4LA, we can only use \ltfla{Many} on expressions that are
defined to be \emph{values} and \emph{use no linear variables}.

@@

I started off with linearity and showing how it helps a programmer keep track
of of memory allocation and deallocation, and then demonstrated how to add
intuitionism for the only values we wish to be intuitionistic. I will now
explain how I enforce safe aliasing and read/write permissions.

Array and matrix types are parameterised by
\emph{fractional-capabilities}~\cite{boyland}.  A fraction of 1 ($2^0$)
represents complete ownership of a value; in particular, this allows a
programmer to write or free it. Creating an array gives you ownership of it;
the function \ltfla{array : !int --o z arr} (where \ltfla{z} represents `0').
Once you have ownership of an array, you can free it: \ltfla{free : z arr --o
unit}.  Importantly, because a linear-value may only be used once, the array
just freed is \emph{out of scope} for following expressions, preventing
use-after-free.  Ownership also enables you to write to the array: \ltfla{set :
z arr --o !int --o !elt --o z arr} (the syntax \ltfla{w[i] := j} is just sugar
for \ltfla{set w i j}). Here, linearity prevents accessing aliases which
represented the array \emph{before} the mutation.

Any fraction less than 1 (for simplicity, limited to $2^{-k}$ in this system,
for a positive integer $k$) represents read-only access. So, the \ltfla{'x}
represents a natural number (either a zero \ltfla{z}, variable \ltfla{'x} or a
successor (+1) of a natural number). Hence, you can read from (index) any array
\ltfla{get : 'x .  'x arr --o !int --o 'x arr * !elt} (the syntax \ltfla{let !v
<- w[i]} is just sugar for \ltfla{let (w, !v) = get _ w i}). In general, a
left-arrow \ltfla{<-} signifies transparent rebinding with returned values: it
means a program can \emph{appear} to use a variable multiple times, important
for keeping LT4LA usable and readable. The underscore is how a programmer tells
the compiler to automatically \emph{infer} the correct fractional-capability
based on the other arguments passed to the function. In conjunction with the
requirement that function declarations need type-annotations for their
arguments, \textbf{this allows a fractional-capability to be \emph{correctly
inferred} in \emph{any} LT4LA expression}.

Fractions exist to provide both safe aliasing and read/write permissions, via
the primitives \ltfla{share : 'x . 'x arr --o ('x s) arr * ('x s) arr} and
\ltfla{unshare: 'x .  ('x s) arr --o ('x s) arr --o 'x arr}.  For the former, two
arrays returned (which happen to just be the given array) can now only be read
from and not written to. If you want to write to this array, you must use the
latter to combine other read-only aliases until you are left with a value of
type \ltfla{z arr}, guaranteeing no other aliases exists.

Given this set-up, a programmer now \emph{statically} has \emph{perfect}
information about aliasing and ownership of values in the program. They can
only write to or free an array only when they own it; ownership guarantees no
other aliases exist. In Figure~\ref{fig:ltfla_kalman}, I show how this perfect
information can be used to write more readable (equational and algebraic) code
using easy to reason about, value-semantic expressions whose \emph{intensional}
behaviour (regarding aliasing, read/write permissions and memory allocation,
re-use and de-allocation)  is also explicitly and precisely described and
\emph{automatically checkable}. Now the programmer need not resort to manually
figuring out and inserting \texttt{noalias} annotations and worrying about what
variables can and cannot be written to, re-used or freed; instead they can let
the loyal and tireless compiler do the heavy lifting.

@@

An if-expression's condition may evaluate either way at run time, so a
programmer must guarantee that both branches use the same set of linear
variables. Writing the type-checker in a pure-functional monadic style paid off
here because I could now sandwich monadic values with state-adjustments either
side of it. Given two monadic values that represented type-checking two
branches of an if-expression, I could use the code in
Figure~\ref{fig:same_resources} to easily save, reset and compare the state
either side of running those monadic values.

@@

A non-recursive function may be used more than once if it does not refer to any
linear variables from the surrounding scope. So, we can desugar something like
\ltfla{let x = 3 in let !f (y : !int) = x + y in <body>} to \ltfla{let x = 3 in
let Many f =  Many (fun y : !int -> x + y) in  <body>}.  Recursion is slightly
more complicated: we can desugar the factorial function
(Figure~\ref{fig:ltfla_factorial}) to \ltfla{let Many f = fix (f, x : !int, if
(*..*) : !int) in <body> }. However, \ltfla{fix}, like \ltfla{Many}, must also
not use any linear variables from its surrounding scope, because calling a
recursive function may evaluate its body multiple times.

\begin{landscape}
    \begin{figure}[p]
    \centering
    \begin{minipage}{1.47\textheight}
    \inputminted[linenos, fontsize=\footnotesize]{fortran}{kalman.f90}
    \end{minipage}
    \caption{Kalman filter in Fortran 90.}\label{fig:fortran_kalman}
    \end{figure}
\end{landscape}

\section{Matrix Expressions}\label{sec:matrix_exps}

We have now arrived at an extended application of linear types. I will show how
we can apply the ability to automatically check aliasing, read/write
permissions, memory allocation, re-use and deallocation, to the domain of
matrix-expression compilation (automatically generating code like
Figure~\ref{fig:fortran_kalman} from mathematical expressions like
Figure~\ref{fig:maths_kalman}).

In Figure~\ref{fig:fortran_kalman}, we see the difficulty of efficiently
implementing a \emph{Kalman filter}~\cite{kalman}, a powerful set of equations
(Figure~\ref{fig:maths_kalman}) applicable to a wide variety of problems. From
the comments, we see that every variable is annotated with the step/matrix
expression that it will hold at some point during the computation (an
equivalent alternative, say in C++, could be to have a meaningful name for each
step and manually keep track of which names alias the same location).

\begin{figure}[tp]
    \begin{align*}
        \mu' &= \mu + \Sigma H^T (R + H \Sigma H^T)^{-1} (H \mu - \textrm{data})\\
        \Sigma' &= \Sigma ( I - H^T (R + H \Sigma H^T)^{-1} H \Sigma )
    \end{align*}
    \caption{Kalman filter equations (credit:
    \href{http://matthewrocklin.com/blog/work/2012/11/24/Kalman-Filter}{matthewrocklin.com}).}\label{fig:maths_kalman}
\end{figure}

In contrast, Figure~\ref{fig:ltfla_kalman}, offers the advantages of
\begin{itemize}
    \item aliasing: labelling each step with a different, more meaningful variable name,
    \item easily spotting which resources are being passed in and which are
        allocated for the function (new/copy),
    \item unambiguously seeing \emph{when} and what values are freed or written-over;
\end{itemize}
and have the compiler automatically ensure the safety of each of the above by respectively
\begin{itemize}
    \item making it impossible to refer to steps/values which are no longer usable,
    \item ensuring all values are declared and \emph{initialised} correctly before they are used,
    \item checking values are neither used after they are freed/written-over \emph{nor} leaked.
\end{itemize}

@@

\begin{figure}[tp]
    \inputminted[linenos, fontsize=\small]{ocaml}{../test/examples/kalman.lt}
    \caption{Kalman filter in LT4LA. In contrast with the Fortran
        (Figure~\ref{fig:fortran_kalman}) or C (Figure~\ref{fig:cblas_kalman})
        implementations, this one is readable (equational and algebraic),
        using easy to reason about, value-semantic expressions. In addition to
        that, its \emph{intensional} behaviour(regarding aliasing, read/write
        permissions and memory allocation, re-use and de-allocation) is also
        explicitly and precisely described and (unlike with Fortran or C)
        \emph{automatically checkable}.}\label{fig:ltfla_kalman}
\end{figure}

@@

All of the syntax in Figure~\ref{fig:ltfla_kalman} can be unambiguously
desugared \emph{before} type-checking, through fairly simple pattern matching.
Matrix expression compilation is well-trodden territory in
academia~\cite{rocklin_thesis, fabregat_thesis, gunnels_flame, linnea, taco}
but this is, to my knowledge, the \textbf{first type-based approach} to it.

@@

\begin{figure}[tp]
    \begin{minted}[fontsize=\small]{ocaml}
let x <- [| y |] in ..              ==> let (y,x) = copyM_to _ y x in ..
let x <- new [| y |] in ..          ==> let (y,x) = copyM _ y in ..
let x <- [| a*b + c |] in ..        ==> let x <- [| 1.*a*b + 1.*c |] in ..
let x <- [| i*a*b + j*c |] in ..    ==> let ((a,b), c) = (* BLAS *) in ..
let x <- new (m, n) [| a*b |] in .. ==> let c = matrix m n in
                                        let x <- [| 1.*a*b + 0.*c |] in ..
    \end{minted}
    \caption{Syntactic translations of matrix expressions to linearly-typed
        matrix functions. Further annotations on the matrix variables
        (\ltfla{sym} or \ltfla{^T}) determine which BLAS routine is called and
        what parameters it is passed.}\label{fig:mat_patterns}

@@

One of the first hurdles I encountered was compositionality: to compile $AB +
C$ it is typically better to use a BLAS routine directly rather than but first
compiling $A$, then $B$, then adding a call to multiply them, then compiling
$C$ and finishing with a call to add the results.

@@

    \item \ltfla{let Many x = Many <exp> in <body>} with \ltfla{let x = <exp> in <body>},

@@

The end result is visible in Figures~\ref{fig:ocaml_sumarray}
and~\ref{fig:ocaml_kalman}:  (surprisingly beautiful) OCaml code that is not
obviously safe, as promised at the start of the chapter.

\begin{figure}[tp]
    \centering
    \begin{minted}[linenos, fontsize=\small]{ocaml}
let rec f i n x0 row =
  if Prim.extract @@ Prim.eqI i n then (row, x0)
  else
    let row, x1 = Prim.get row i in
    f (Prim.addI i (Many 1)) n (Prim.addE x0 x1) row
in
f
    \end{minted}
    \caption{Recursive OCaml function for a summing an array, generated (at
        \emph{compile time}) from the code in Figure~\ref{fig:ltfla_sumarray},
        passed through \texttt{ocamlformat} for presentation.}\label{fig:ocaml_sumarray}

@@

\begin{figure}[p]
    \begin{minted}[linenos, fontsize=\footnotesize]{ocaml}
let kalman sigma h mu r_1 data_1 =
  let h, _p_k_n_p_ = Prim.size_mat h in
  let k, n = _p_k_n_p_ in
  let sigma_h = Prim.matrix k n in
  let (sigma, h), sigma_h =
    Prim.symm (Many true) (Many 1.) sigma h (Many 0.) sigma_h
  in
  let (sigma_h, h), r_2 =
    Prim.gemm (Many 1.) (sigma_h, Many false) (h, Many true) (Many 1.) r_1
  in
  let (h, mu), data_2 =
    Prim.gemm (Many 1.) (h, Many false) (mu, Many false) (Many (-1.)) data_1
  in
  let h, new_h = Prim.copy_mat_to h sigma_h in
  let r_2, new_r = Prim.copy_mat r_2 in
  let chol_r, sol_h = Prim.posv new_r new_h in
  let chol_r, sol_data = Prim.potrs chol_r data_2 in
  let () = Prim.free_mat chol_r in
  let h_sol_h = Prim.matrix n n in
  let (h, sol_h), h_sol_h =
    Prim.gemm (Many 1.) (h, Many true) (sol_h, Many false) (Many 0.) h_sol_h
  in
  let () = Prim.free_mat sol_h in
  let h_sol_data = Prim.matrix n (Many 1) in
  let (h, sol_data), h_sol_data =
    Prim.gemm (Many 1.) (h, Many true) (sol_data, Many false) (Many 0.) h_sol_data
  in
  let mu, mu_copy = Prim.copy_mat mu in
  let (sigma, h_sol_data), new_mu =
    Prim.symm (Many false) (Many 1.) sigma h_sol_data (Many 1.) mu_copy
  in
  let () = Prim.free_mat h_sol_data in
  let h_sol_h_sigma = Prim.matrix n n in
  let (sigma, h_sol_h), h_sol_h_sigma =
    Prim.symm (Many true) (Many 1.) sigma h_sol_h (Many 0.) h_sol_h_sigma
  in
  let sigma, sigma_copy = Prim.copy_mat_to sigma h_sol_h in
  let (sigma, h_sol_h_sigma), new_sigma =
    Prim.symm (Many false) (Many (-1.)) sigma h_sol_h_sigma (Many 1.) sigma_copy
  in
  let () = Prim.free_mat h_sol_h_sigma in
  ((sigma, (h, (mu, (r_2, sol_data)))), (new_mu, new_sigma)) )
in
kalman
    \end{minted}
    \caption{OCaml code for a Kalman filter, generated (at \emph{compile time})
        from the code in Figure~\ref{fig:ltfla_kalman}, passed through
        \texttt{ocamlformat} for presentation.}\label{fig:ocaml_kalman}

@@

Advantages of generating code at compile time include avoiding runtime overhead
as well as catching any interface changes (type-signatures now include
\emph{intensional} information about which parameters are read from or written
to) between the generated code and the code that uses it via OCaml's type
system, \emph{another benefit} (in addition to a sanity check on LT4LA's
type-checker implementation) of \emph{embedding} LT4LA's type system inside of
OCaml's.

I suspect that this approach will be very valuable to not just \emph{users} of
libraries such as Numpy or Owl, but also their \emph{implementors}: library
functions which present a safe, value-semantic interface but use unsafe,
mutating operations on the inside could now be expressed using LT4LA and have
an extra safety-net and automatic checking for their implementations.
